---
title: "Find DMR using coMethDMR"
author: 
  - Wei Zhang^[University of Miami]
  - Lily Wang^[University of Miami]
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    highlight: pygments
    theme: yeti
    toc: true
    number_sections: true
    df_print: paged
    code_download: false
    toc_float:
      collapsed: yes
    toc_depth: 3
editor_options:
  chunk_output_type: inline    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = F)

dir.base <- "~/TBL Dropbox/Wei Zhang/AD-Aging-blood-sample-analysis"
# Create and retrieve dir
dir.data <- file.path(dir.base,"../DATASETS/")
dir.data.dmr <- file.path(dir.base, "datasets/DMR/")
dir.data.aux <- file.path(dir.data, "Aux_Sync")

# Create results dir
dir.results <- file.path(dir.base, "analysis_results/DMR")
dir.results.meta <- file.path(dir.base, "analysis_results/meta_analysis")
#for(p in grep("dir.",ls(),value = T)) dir.create(get(p),recursive = TRUE,showWarnings = FALSE)
library(tidyverse)
```

```{r warning=F,message=F, eval = T}
# Required packages
library(coMethDMR)
library(BiocParallel)
library(SummarizedExperiment)
library(survival)
library(janitor)
# Required functions
source(file.path(dir.base, "code/utility/meta.R"))
source(file.path(dir.base, "code/utility/coMethDMR_aux.R"))
source(file.path(dir.base, "code/utility/annotation_and_bacon.R"))
```

# Extract clusters of CpGs located closely 

```{r eval = F}
region <- get_genomic_range(
  array = "EPICv1",
  genome = "hg38",
  dir.data.aux = dir.data.aux
)

WriteCloseByAllRegions(
  fileName = file.path(dir.data.dmr, "EPIC_hg38_region.rds"),
  region = region,
  genome = "hg38",
  arrayType = "EPIC",
  maxGap = 200,
  minCpGs = 3
)
```

```{r}
coMeth_region <- readRDS(file.path(dir.data.dmr, "EPIC_hg38_region.rds"))
```

# Find ADNI DMR

## Load Data

```{r}
dir.data <- file.path(dir.base, "../DNAm-Longitudinal-Cognition/datasets/ADNI/BEclear") 
adni_se <- readRDS(
  file.path(dir.data, "ADNI_BEclear_se.rds")
)
# Remove samples < 65
adni_se <- adni_se[,adni_se$age_at_visit >= 65] # 211
# Remove DEM_STATUS == 1
adni_se <- adni_se[,adni_se$DEM_STATUS == 0] # 193
pheno_adni <- colData(adni_se) %>%
  data.frame()
beta <- assay(adni_se)
```

## Find comethylated clusters

```{r}
celltype <- c("B", "NK", "CD4T", "Mono", "Gran")
resid_mat <- getResid(
  mat = beta,
  betaToM = T,
  pheno = pheno_adni %>% dplyr::select(c("PTGENDER", celltype)),
  nCore = 50
)
### Find co-methylated clusters
coMeth_ls <- CoMethAllRegions(
  dnam = resid_mat,      
  betaToM = F,                   
  CpGs_ls = coMeth_region,
  genome = "hg38",
  arrayType = "EPIC",
  method = "spearman",             
  returnAllCpGs = TRUE,              
  output = "CpGs",
  nCores_int = 50,
  progressbar = TRUE
)
save(
  resid_mat,
  coMeth_ls,
  file = file.path(dir.data.dmr,"ADNI_resid_mat_and_coMeth_ls.rda")
)
```

## Map input region to output region

```{r}
load(file.path(dir.data.dmr,"ADNI_resid_mat_and_coMeth_ls.rda"))
```

```{r}
input_region <- names(coMeth_region)
output_region <- names(coMeth_ls)
map_table <- map_input_output_region(input_region, output_region)
```

## Test with regression on median value

```{r}
M <- minfi::logit2(beta)
coMeth.results <- getSigRegion(
  coMeth.list = coMeth_ls,
  M = M,
  pheno = pheno_adni,
  test_var = "age_at_visit",
  adjust_var = c("PTGENDER", celltype),
  nCore = 30
)
coMeth.results <- coMeth.results %>% 
  dplyr::mutate(input_region = map_table$input_region[match(region, map_table$output_region)],
                out_region = region, .keep = "unused", .before = 1) 
```

```{r}
## For multiple region, select the smallest pvalue
coMeth.results <- coMeth.results %>% 
  group_by(input_region) %>% 
  slice_min(pr_t, with_ties = F)
```

```{r}
coMeth.results %>% readr::write_csv(
  file.path(dir.results, "ADNI_coMeth_median_results.csv")
)
```

## Bacon correction

```{r warning = F}
coMeth.results$z <- qnorm(coMeth.results$pr_t)
result_bacon <- bacon_adj(
  coMeth.results,
  est_var = "estimate",
  z_var = "z",
  std_var = "std_error",
  save = T,
  dir.save = dir.results,
  prefix = "ADNI_coMeth_median")
```

```{r echo = F, eval = T}
readxl::read_xlsx(
  file.path(dir.results, "ADNI_coMeth_median_inflation_stats.xlsx")
) %>% 
  t() %>%
  data.frame(values = .)
```


# Find FHS DMR

## Load Data

```{r}
# Load data from previous analysis
dir.data <- file.path(dir.base,"../DATASETS/Framingham")
dir.data.processed <- file.path(dir.data, "DNAm/processed/EPIC") 
framingham_se <- readRDS(
  file.path(dir.data.processed, "FHS_OffSpring_EPIC_rm_dem_nw_se.RDS")
)
# Remove samples < 65
framingham_se <- framingham_se[,framingham_se$Age_at_collection >= 65] # 741
# Remove DEM_STATUS == 1
framingham_se <- framingham_se[,framingham_se$DEM_STATUS == 0] # 699
pheno <- colData(framingham_se) %>%
  data.frame()
beta <- assay(framingham_se)
# Select samples that have highest bc score in each pedno
pheno_fhs <- pheno %>% 
  group_by(pedno) %>% 
  slice_max(bisulfiteConversion) #282
beta_selected <- beta[,pheno_fhs$sample]
```

## Find comethylated clusters

```{r}
celltype <- c("B", "NK", "CD4T", "Mono", "Gran")
resid_mat <- getResid(
  mat = beta_selected,
  betaToM = T,
  pheno = pheno_fhs %>% dplyr::select(c("sex", celltype)),
  nCore = 50
)
### Find co-methylated clusters
coMeth_ls <- CoMethAllRegions(
  dnam = resid_mat,      
  betaToM = F,                   
  CpGs_ls = coMeth_region,
  arrayType = "EPIC",
  genome = "hg38",
  method = "spearman",             
  returnAllCpGs = TRUE,              
  output = "CpGs",
  nCores_int = 50,
  progressbar = TRUE
)
save(
  resid_mat,
  coMeth_ls,
  file = file.path(dir.data.dmr,"FHS_resid_mat_and_coMeth_ls_combp.rda")
)
```

## Map input region to output region

```{r}
load(file.path(dir.data.dmr,"FHS_resid_mat_and_coMeth_ls_combp.rda"))
```

```{r}
input_region <- names(coMeth_region)
output_region <- names(coMeth_ls)
map_table <- map_input_output_region(input_region, output_region)
```

## Test with survival

```{r}
M <- minfi::logit2(beta_selected)
coMeth.results <- getSigRegion(
  coMeth.list = coMeth_ls,
  M = M,
  pheno = pheno_fhs,
  test_var = "Age_at_collection",
  adjust_var = c("sex", celltype),
  nCore = 50
)
coMeth.results <- coMeth.results %>% 
  dplyr::mutate(input_region = map_table$input_region[match(region, map_table$output_region)],
                out_region = region, .keep = "unused", .before = 1) 
```

```{r}
## For multiple region, select the smallest pvalue
coMeth.results <- coMeth.results %>% 
  group_by(input_region) %>% 
  slice_min(pr_t, with_ties = F)
```

```{r}
coMeth.results %>% readr::write_csv(
  file.path(dir.results, "FHS_coMeth_median_results.csv")
)
```

## Bacon correction

```{r warning = F}
coMeth.results$z <- qnorm(coMeth.results$pr_t)
result_bacon <- bacon_adj(
  coMeth.results,
  est_var = "estimate",
  z_var = "z",
  std_var = "std_error",
  save = T,
  dir.save = dir.results,
  prefix = "FHS_coMeth_median")
```

```{r echo = F, eval = T}
readxl::read_xlsx(
  file.path(dir.results, "FHS_coMeth_median_inflation_stats.xlsx")
) %>% 
  t() %>%
  data.frame(values = .)
```

# Meta analysis

## Analysis

```{r}
adni_results <- read_csv(file.path(dir.results, "ADNI_coMeth_median_bacon_correction.csv"))
results_adni <- adni_results %>% 
  mutate(study = "ADNI",
         study_id = "ADNI")
fhs_results <- read_csv(file.path(dir.results, "FHS_coMeth_median_bacon_correction.csv"))
results_fhs9 <- fhs_results %>% 
  mutate(study = "FHS",
         study_id = "FHS9")
```

```{r}
# Create list for meta analysis
results_list <- list(
  FHS9 = results_fhs9,
  ADNI = results_adni
)
# Function meta_wrapper load from meta.R
results <- meta_wrapper(
  results_list = results_list,
  effect = "Estimate.bacon",
  se = "StdErr.bacon",
  sm = "MD",
  test_var = "input_region",
  select_var = NULL
)
# Save results
write_csv(
  results,
  file.path(dir.results,
            "meta_analysis_FHS9_ADNI_DMR_results.csv")
)
```

## Annotation

```{r}
results <- read_csv(
  file.path(dir.results,
            "meta_analysis_FHS9_ADNI_DMR_results.csv"),
  show_col_types = F
)
results_meta <- read_csv(
  file.path(dir.results.meta, "meta_analysis_FHS9_ADNI_results.csv"),
  show_col_types = F
)
```

```{r}
results_anno <- add_dmr_annotation(
  result = results,
  cpg = results_meta$cpg,
  dir.data.aux = dir.data.aux,
  region_var = "input_region",
  array = "EPIC",
  genome = "hg38"
)
```

```{r}
write_csv(
  results_anno,
  file.path(dir.results, "meta_analysis_FHS9_ADNI_DMR_results_annotated.csv")
)
```

## Significant results

```{r eval = T, echo = F}
results_anno <- read_csv(
  file.path(dir.results, "meta_analysis_FHS9_ADNI_DMR_results_annotated.csv"),
  show_col_types = F
)
results_anno %>% 
  filter(fdr < 0.05 & direction %in% c("++", "--"))
```

```{r}
write_csv(
  results_anno %>% 
  filter(fdr < 0.05 & direction %in% c("++", "--")),
  file.path(dir.results, "meta_analysis_FHS9_ADNI_DMR_results_annotated_filtered.csv")
)
```

# Merge coMethDMR with combp results

```{r eval = T}
combp_results <- read_csv(
  file.path(dir.results, "combp_filtered_results_annotated.csv"),
  show_col_types = F
)
coMeth_results <- read_csv(
  file.path(dir.results, "meta_analysis_FHS9_ADNI_DMR_results_annotated_filtered.csv"),
  show_col_types = F
)
```

## Combp significant regions

```{r echo = F, eval = T}
combp_results
```

## Find overlap region

```{r  eval = T}
coMeth_region <- coMeth_results$input_region
combp_region <- combp_results$region
overlapping_rg <- map_overlap_region(coMeth_region, combp_region)
```

```{r eval = T}
coMeth_region <- coMeth_region[-overlapping_rg$output_idx]
coMeth_region <- c(coMeth_region, overlapping_rg$output_region)
ggvenn::ggvenn(list("Sig. regions in combp" = combp_region,
                    "Sig. regions in coMethDMR" = coMeth_region),
               text_size = 4,
               set_name_size = 4)
```

```{r}
ggsave(file.path(dir.base, "analysis_results/plots/venn_plot_sig_regions.pdf"),
       width = 6, height = 4)
```

## Select overlap region

```{r}
coMeth_results_overlap <- coMeth_results[overlapping_rg$input_idx,]
combp_results_overlap <- combp_results[overlapping_rg$output_idx,]
```

```{r}
combp_df <- combp_results_overlap %>% 
  dplyr::select(
    DMR = region, num_probes, z_p, z_sidak_p, direction, cpgs_in_region,
    GREAT_annotation, UCSC_RefGene_Group, UCSC_RefGene_Name, Relation_to_Island, E073_15_coreMarks_segments_state
  )
dmr_df <- coMeth_results_overlap %>% 
  dplyr::select(
    Rgion = input_region, estimate, se, pVal.fixed, fdr, direction_cohort = direction,
    ADNI_region = ADNI_out_region, ADNI_n, ADNI_Estimate.bacon, ADNI_StdErr.bacon, ADNI_pValue.bacon, 
    FHS_region = FHS9_out_region, FHS9_n, FHS9_Estimate.bacon, FHS9_StdErr.bacon, FHS9_pValue.bacon
  )
```

## Save results

```{r}
writexl::write_xlsx(
  list(coMethDMR = coMeth_results_overlap,
       combp = combp_results_overlap),
  file.path(dir.results, "coMeth_combp_overlap_dmr.xlsx")
)
```

```{r}
write_csv(
  cbind(combp_df, dmr_df) %>% arrange(z_p),
  file.path(dir.results, "coMeth_combp_overlap_dmr.csv")
)
```

## Combine with meta-analysis results

```{r}
meta_results <- read_csv(
  file.path(dir.results.meta,
            "meta_analysis_FHS9_ADNI_results.csv")
)
```

```{r}
# top 10 dmr
## combp
combp_df_top10 <- combp_results %>% 
  slice_min(z_p, n = 10) %>%
  arrange(z_p) %>%
  separate_longer_delim(cpgs_in_region, ",") %>% 
  dplyr::select(region, cpg = cpgs_in_region, z_p, z_sidak_p) 
combp_df_top10 <- left_join(combp_df_top10, meta_results)

## ADNI
coMethDMR_top10_adni <- coMeth_results %>% 
  slice_min(ADNI_pValue.bacon, n = 10) %>% 
  arrange(ADNI_pValue.bacon) %>%
  separate_longer_delim(ADNI_CpGsinRegion, ",") %>% 
  dplyr::select(ADNI_region = ADNI_out_region, cpg = ADNI_CpGsinRegion, 
                ADNI_coMeth_estimate = ADNI_Estimate.bacon,
                ADNI_coMeth_se = ADNI_StdErr.bacon,
                ADNI_coMeth_pval = ADNI_pValue.bacon,
                ADMO_coMeth_fdr = ADNI_fdr.bacon) 
coMethDMR_top10_adni <- left_join(
  coMethDMR_top10_adni, 
  meta_results %>% dplyr::select(
    cpg, seqnames, start, end, Relation_to_UCSC_CpG_Island, UCSC_RefGene_Name, 
    UCSC_RefGene_Group, GREAT_annotation, 
    ADNI_single_cpg_estimate = ADNI_Estimate.bacon, 
    ADNI_single_cpg_se = ADNI_StdErr.bacon,
    ADNI_single_cpg_pval = ADNI_pValue.bacon,
    ADNI_single_cpg_fdr = ADNI_fdr.bacon
  )
)

## FHS
coMethDMR_top10_fhs <- coMeth_results %>% 
  slice_min(FHS9_pValue.bacon, n = 10) %>% 
  arrange(FHS9_pValue.bacon) %>%
  separate_longer_delim(FHS9_CpGsinRegion, ",") %>% 
  dplyr::select(FHS_region = FHS9_out_region, cpg = FHS9_CpGsinRegion, 
                FHS_coMeth_estimate = FHS9_Estimate.bacon,
                FHS_coMeth_se = FHS9_StdErr.bacon,
                FHS_coMeth_pval = FHS9_pValue.bacon,
                FHS_coMeth_fdr = FHS9_fdr.bacon) 
coMethDMR_top10_fhs <- left_join(
  coMethDMR_top10_fhs, 
  meta_results %>% dplyr::select(
    cpg, seqnames, start, end, Relation_to_UCSC_CpG_Island, UCSC_RefGene_Name, 
    UCSC_RefGene_Group, GREAT_annotation, 
    FHS_single_cpg_estimate = FHS9_Estimate.bacon, 
    FHS_single_cpg_se = FHS9_StdErr.bacon,
    FHS_single_cpg_pval = FHS9_pValue.bacon,
    FHS_single_cpg_fdr = FHS9_fdr.bacon
  )
)

writexl::write_xlsx(
  list(combp_top10 = combp_df_top10,
       coMeth_ADNI_top10 = coMethDMR_top10_adni,
       coMeth_FHS_top10 = coMethDMR_top10_fhs),
  file.path(dir.results, "coMeth_combp_overlap_dmr_top10.xlsx")
)
```

# Session Information

<details>
  <summary>**Session Info**</summary>
```{r session, eval = T}
sessionInfo()
```
</details>
